%\section{Introduction}
%\label{sec:intro}

The \emph{Syntax-Guided Synthesis Competition} (SyGuS-Comp) is an annual competition aimed to provide an objective platform for comparing different approaches for solving the \emph{Syntax-Guided Synthesis} (SyGuS) problem. A SyGuS problem takes as input a logical specification $\varphi$ for what a synthesized function $f$ should compute, and a grammar $G$ providing syntactic restrictions on the function $f$ to be synthesized. Formally, a solution to a SyGuS instance $(\varphi,G,f)$ is a function $f_{imp}$ that is expressible in the grammar $G$ such that the formula $\varphi[f/f_{imp}]$ obtained by replacing  $f$ by $f_{imp}$ in the logical specification $\varphi$ is valid. SyGuS instances are formulated in SyGuS-IF~\cite{RaghothamanU14}, a format built on top of SMT-LIB2~\cite{smtlib}.

We report here on the 4th SyGuS competition that took place in July 2017, in Heidelberg, Germany as a satellite event of CAV'17 (The 29th International Conference on Computer Aided Verification) and SYNT'17 (The Sixth Workshop on Synthesis). As in the previous competition there were four tracks: the general track, the conditional linear integer arithmetic track, the invariant synthesis track, and the programming by examples track. We assume most readers of this report are already familiar with the SyGuS problem and the tracks of SyGuS-Comp and thus refer the unfamiliar reader to the report on last year's competition~\cite{SyGuSComp15}. 

The report is organized as follows. Section~\ref{sec:benchs} describes the participating benchmarks. Section~\ref{sec:solvers}  lists the participating solvers, and briefly describes the main idea behind their strategy.  Section~\ref{sec:exp-set} provides details on the experimental setup. Section~\ref{sec:comp-results} gives an overview of the results per track. Section~\ref{sec:benchs-pres} provides details on the results, given from a single benchmark respective. Section~\ref{sec:discussion} concludes.